# Functional enrichment analysis

Here we perform gene ontology, KEGG pathway and Reactome pathway enrichment using the
[clusterProfiler](https://bioconductor.org/packages/release/bioc/vignettes/clusterProfiler/inst/doc/clusterProfiler.html)
package with some custom plotting functions. We can perform overrepresentation analysis (overrep) 
or gene set enrichment analysis (gsea).

**Analysis mode:** `r analysis.mode`

```{r cprof_setup, results='hide'}

# output directory for clusterProfiler results
cprof.folder <- "clusterprofiler"
plots.folder <- "clusterprofiler/plots"

# check for subfolder and create if absent
ifelse(!dir.exists(file.path('.',cprof.folder)),
       dir.create(file.path('.',cprof.folder), recursive = TRUE), NA)
ifelse(!dir.exists(file.path('.',plots.folder)),
       dir.create(file.path('.',plots.folder), recursive = TRUE), NA)

# NOTE: keyTypes to obtain from bitr-------------------------------------------
types <- c("ENTREZID","SYMBOL","UNIPROT")

# NOTE: convert IDs to symbols for KEGG?---------------------------------------
id.convert <- TRUE

```

```{r bitr_genes, results='hide'}
# NOTE: Gene IDs are provided in what format?----------------------------------
from.type <- "FLYBASE"

deg.list <- nested.lapply(sel.list, rownames)
genes.df <- nested.lapply(deg.list, bitr, fromType=from.type, toType=types, OrgDb=orgdb)
```

```{r enrich_all, results='hide'}

# run enrichment for all gene lists over a loop

# NOTE: settings for clusterProfiler and plots---------------------------------
go.keytype <- 'ENTREZID'
kegg.keytype <- 'ncbi-geneid'
pvalueCutoff <- 1
qvalueCutoff <- 1
pAdjustMethod <- 'BH'
go.uni <- NULL
path.uni <- NULL
kegg.organism <- 'dme'
RUN.REACTOME <- FALSE
reactome.organism <- 'fly'

# list to hold all enrichment results
# will have structure:
#
#   all.enrich[[contrast]][[selection]][[enrichment label]]
#
all.enrich <- list()
enrich.files <- list()
# loop over each contrast
for(comp in names(genes.df)){

    all.enrich[[comp]] <- list()
    enrich.files[[comp]] <- list()

    # loop over selection list
    for(sel in sel.names){
        all.enrich[[comp]][[sel]] <- list()
        enrich.files[[comp]][[sel]] <- list()

        if(analysis.mode == 'gsea'){
            # for 'gsea' we create a decreasing ranked named vector of values

            # named vector with ID -> entrezID mapping
            idmap <- genes.df[[comp]][[sel]][[go.keytype]]
            names(idmap) <- genes.df[[comp]][[sel]][[from.type]]
    
            # we use the 'stat' column as the ranking metric
            gene <- sel.list[[comp]][[sel]]$stat
            names(gene) <- idmap[rownames(sel.list[[comp]][[sel]])]
            gene <- gene[order(gene, decreasing=TRUE)]

            # removing NA's
            gene <- gene[!is.na(gene)]
            gene <- gene[!is.na(names(gene))]
        } else if(analysis.mode == 'overrep'){
            # for 'overrep' we supply the gene list
            gene <- genes.df[[comp]][[sel]][[go.keytype]]
        }

        # perform GO analysis
        for(ont in c('CC', 'BP', 'MF')){
            res.label <- paste('GO', ont, analysis.mode, sep='_')

            if(analysis.mode == 'gsea'){
                go.res <- gseGO(
                    geneList=gene, keyType=go.keytype, OrgDb=orgdb,
                    ont=ont, pvalueCutoff=pvalueCutoff, pAdjustMethod=pAdjustMethod)
            } else if(analysis.mode == 'overrep'){
                go.res <- enrichGO(
                    gene=gene, universe=go.uni, keyType=go.keytype, OrgDb=orgdb,
                    ont=ont, pvalueCutoff=pvalueCutoff, qvalueCutoff=qvalueCutoff,
                    readable = TRUE)
            }

            all.enrich[[comp]][[sel]][[ont]] <- go.res
            if (!is.null(go.res)){
                enrich.files[[comp]][[sel]][[ont]] <- write.clusterprofiler.results(go.res, cprof.folder,
                                                        paste0(res.label, '_', comp, '_', sel),
                                                        analysis.mode=analysis.mode)
            }
        }

        # perform KEGG analysis
        res.label <- paste('KEGG', analysis.mode, sep='_')
        if(analysis.mode == 'gsea'){
            kegg.res <- gseKEGG(
                geneList=gene, organism=kegg.organism,
                keyType=kegg.keytype, pvalueCutoff=pvalueCutoff,
                pAdjustMethod=pAdjustMethod)
        } else if(analysis.mode == 'overrep'){
            kegg.res <- enrichKEGG(
                gene=gene, universe=path.uni, organism=kegg.organism,
                keyType=kegg.keytype, pvalueCutoff=pvalueCutoff,
                qvalueCutoff=qvalueCutoff)

        }

        all.enrich[[comp]][[sel]][['kegg']] <- kegg.res
        if(!is.null(kegg.res) && nrow(kegg.res) > 0){
            enrich.files[[comp]][[sel]][['kegg']] <- write.clusterprofiler.results(kegg.res, cprof.folder,
                                                    paste0(res.label, '_', comp, '_', sel),
                                                    analysis.mode=analysis.mode)
        }

        if (RUN.REACTOME) {
            # NOTE: Install OrgDb--------------------------------------------------
            #   ReactomePA requires the relevant OrgDb package to be installed
            #   and so is disabled by default.
            res.label <- paste('Reactome', analysis.mode, sep='_')
            if(analysis.mode == 'overrep'){
                reactome.res <-enrichPathway(
                    gene=gene, universe=path.uni, organism=reactome.organism,
                    pvalueCutoff=pvalueCutoff, qvalueCutoff=qvalueCutoff, readable=TRUE)
            } else if(analysis.mode == 'gsea'){
                reactome.res <- gsePathway(
                    gene=gene, organism=reactome.organism,
                    pvalueCutoff=pvalueCutoff, pAdjustMethod=pAdjustMethod)
            }
            all.enrich[[comp]][[sel]][['reactome']] <- reactome.res

            if (!is.null(reactome.res)){
                enrich.files[[comp]][[sel]][['reactome']] <- write.clusterprofiler.results(reactome.res, cprof.folder,
                                                                paste0(res.label, '_', comp, '_', sel),
                                                                analysis.mode=analysis.mode)
            }
        }
    }
}

# convert IDs to symbols for readable plots
# for 'gsea' this conversion is done for all pathway results
# for 'overrep' only kegg result needs conversion
if(id.convert){
    for(comp in names(all.enrich)){
        for(sel in names(all.enrich[[comp]])){
            all.go <- names(all.enrich[[comp]][[sel]])
    
            # for 'overrep' do conversion only for KEGG
            if(analysis.mode == 'overrep'){
                if('kegg' %in% all.go){
                    all.go <- c('kegg')
                    id.col <- 'geneID'
                } else {
                    next
                }
            } else if(analysis.mode == 'gsea'){
                id.col <- 'core_enrichment'
            }
    
            for(go in all.go){
                res <- all.enrich[[comp]][[sel]][[go]]
                if(!is.null(res) && nrow(res) > 0){
                    res <- set.readable(res, genes.df=genes.df, id.col=id.col)
                    all.enrich[[comp]][[sel]][[go]] <- res
                }
            }
        }
    }
}
                
```

```{r adjust_catlen}

# This shortens the names of functional terms for better visualization
maxcatlen <- 50
for(comp in names(all.enrich)){
  for(name in names(all.enrich[[comp]])){
    for(go in names(all.enrich[[comp]][[name]])){
            all.enrich[[comp]][[name]][[go]]@result$Description <- substr(all.enrich[[comp]][[name]][[go]]@result$Description, 1, maxcatlen)
    }
  }
}

```

Below we show the results separately for each comparison. Links to files
containing analysis results are shown above each set of plots, with parsed
files showing genes on separate lines.

```{r plot_go, fig.width=15, fig.height=10, results='asis'}
# go labels for file names
go.label <- list(
    BP='GO Biological Process',
    CC='GO Cellular Component',
    MF='GO Molecular Function',
    kegg='KEGG Pathways',
    reactome='Reactome Pathways')

comp.label <- lapply(res.list, function (x) x[['label']])

for(comp in names(all.enrich)){
    mdcat('## ', comp.label[[comp]], ' {.tabset}')
    for(go in names(go.label)){

        mdcat('### ', go.label[[go]], ' {.tabset}')

        mdcat('TSV format results (condensed and split versions):\n')

        dotplots <- list()
        emapplots <- list()
        cnetplots <- list()

        for (sel in sel.names){
            res <- all.enrich[[comp]][[sel]][[go]]

            if (is.null(res) || nrow(res) == 0)
            {
                cat('\n\n\nNo genes enriched for:', comp, sel, go, '\n\n\n')
                next
            }

            outfiles <- enrich.files[[comp]][[sel]][[go]]
            f1 <- outfiles[['orig']]
            f2 <- outfiles[['split']]
            mdcat(' - ', sel, ' [', f1, '](', f1 ,')', ', [', f2, '](', f2 ,')')
            title <- paste(sel, go)

            # NOTE: arguments to dotplot, emapplot, and cnetplot---------------
            #   Arguments provided are package defaults, explicitly provided
            #   here for better visibility on what to tweak. Most commonly
            #   adjusted is probably the showCategory argument.

            dotplots[[sel]] <- dotplot(
                res,
                showCategory=10,
                color='p.adjust',
                x='GeneRatio'
            ) +
                ggtitle(title) +
                theme(plot.title=element_text(hjust=0.5, size=15, face='bold'))

            emapplots[[sel]] <- emapplot(
                res,
                showCategory=30,
                color='p.adjust'
            ) +
                ggtitle(title) +
                theme(plot.title=element_text(hjust=0.5, size=15, face='bold'))

            cnetplots[[sel]] <- cnetplot(
                res,
                showCategory=5,
                foldChange=NULL,
                colorEdge=FALSE,
                circular=FALSE,
                node_label="all"
            ) +
                ggtitle(title) +
                theme(plot.title=element_text(hjust=0.5, size=15, face='bold'))
        }

        save.and.link <- function(plottype){
            f <- file.path(plots.folder, paste0(go,"_", comp,"_", plottype, ".pdf"))
            dev.copy(pdf, file=f, width=15, height=10)
            dev.off()
            mdcat('Link to pdf - [', f, '](', f, ')')
            cat('\n')
        }

        # It's possible that we got this far without any results to plot; if so
        # then skip creating the tabs and plots
        skip <- (length(dotplots) == 0) & (length(emapplots) == 0) & (length(cnetplots) == 0)


        if (!skip){
            mdcat('#### Dotplot')
            print(plot_grid(plotlist=dotplots, align='hv'))
            save.and.link('dotplot')

            mdcat('#### Emapplot')
            print(plot_grid(plotlist=emapplots, align='hv'))
            save.and.link('emapplot')

            mdcat('#### Cnetplot')
            print(plot_grid(plotlist=cnetplots, align='hv'))
            save.and.link('cnetplot')
        }
    }
}
```
