# Gene patterns {.tabset}

We can roughly group genes into expression patterns. This uses the [DEGreport
package](https://www.bioconductor.org/packages/release/bioc/html/DEGreport.html),
which in turn uses the
[ConsensusClusterPlus](https://www.bioconductor.org/packages/release/bioc/html/ConsensusClusterPlus.html)
algorithm to cluster genes into similar expression patterns. The lists of genes
found in each cluster are reported below the plot. Cluster IDs are also reported in
the exported results TSVs in the section below.

The default general flow is as follows:

```{r dpsettings}
# Settings for below chunks
# Docs: https://lcdb.github.io/lcdb-wf/rnaseq-rmd.html#dpsettings

# NOTE: settings for DEGpatterns-----------------------------------------------
#  These parameters are set up here because we will be providing them to
#  multiple degPatterns calls.

# NOTE:  Max number of genes to include----------------------------------------
#   If more genes are differeentially expressed, they will be downsampled to
#   this many
lim <- 2000


# NOTE: Min number of genes to include----------------------------------------
#   If fewer than this many genes, skip the clustering
# lower.lim <- 0  # [ TEST SETTINGS +1 ]
lower.lim <- 10

# NOTE:  Correlation coefficient above which to merge clusters-----------------
cutoff <- 0.5

# NOTE: x-axis for degPatterns ------------------------------------------------
#   Should be a column in the colData metadata
time <- "group"

# NOTE: color for degPatterns.------------------------------------------------
#   Use NULL for no color.
col <- NULL

# NOTE: This is set very low for test data. Default is 15.---------------------
#   Minimum cluster size.
# minc <- 1  # [ TEST SETTINGS +1 ]
minc <- 15

# NOTE: This is a very low value used for getting the degPatterns to run -----
low.minc <- 1
```

- Take the union of all genes that are changed across all contrasts
- If there are more than `r lim` genes, randomly sample `r lim` genes (to keep the computations reasonable)
- Cluster genes by coexpression across contrasts
- If clusters are correlated with a correlation coefficient of `r cutoff` or
  more, they are merged together
- Clusters with fewer than `r minc` genes are not shown.

```{r finalclusters, fig.width=12, results='asis', cache=TRUE, dependson='selections'}
# Run the clustering, identify patterns, and generate plots.
# Docs: https://lcdb.github.io/lcdb-wf/rnaseq-rmd.html#finalclusters
# NOTE: which genes to cluster?------------------------------------------------
#    By default, we get all the changed genes, but you may want only the up or
#    down genes.
ll <- lapply(res.list, function (x) get.sig(x[['res']], 'changed'))

# Filter out results where there were zero genes detected.
ll <- ll[lapply(ll, length) > 0]

add.cluster.id <- function(clusters, res, label){
    # Merges the degPattern cluster IDs `cluster` with DESeqresults `res`
    # `label` will be used to create a cluster column with a unique column name 
    # returns a DESeqresults with cluster IDs
    unq <- unique(clusters[, c('genes', 'cluster')])
    names(unq)[names(unq) == 'genes'] <- 'gene'
    merged <- merge(as.data.frame(res) %>% tibble::rowid_to_column("ID"),
                    unq, by='gene', all.x=TRUE) %>%
        arrange(ID)
    res[[paste0(label, '_cluster')]] <- merged[['cluster']]
    return(res)
}

for (name in names(ll)){
    # Print a nice Markdown header
    mdcat('## ', res.list[[name]][['label']])

    genes <- ll[[name]]

    # Plotting a large number of genes will take a lot of time to perform the
    # clustering. If there are more than this limit, then we take a random
    # sample.
    if (length(genes) > lim){
        genes <- sample(genes, lim)
    }

    if (length(genes) < lower.lim){
        mdcat('**Too few genes (', length(genes), ') to cluster**')
        next
    }

    # Extract the normalized counts for these genes
    vsd.i <- varianceStabilizingTransformation(dds.list[[ res.list[[name]][['dds']] ]], blind=TRUE)
    idx <- rownames(vsd.i) %in% genes
    ma <- assay(vsd.i)[idx,]

    # Remove genes with identical normalized counts across all samples
    ma <- ma[apply(as.data.frame(ma), 1, n_distinct) > 1, ]

    colData.i <- colData(dds.list[[ res.list[[name]][['dds']] ]])
    colData.i <- colData.i[,!(colnames(colData.i) %in% exclude.for.printing)]


    # Sometimes, if there are limited clusters, degPattern fails. The solution
    # below is to reduce the `minc` to a very low number which allows it to run
    # without error, disable plotting, then manually remove the clusters with
    # low numbers of genes, then use the degPlotCluster function to re-plot
    # them.
    p <- degPatterns(
        ma,
        colData.i,
        time=time,
        col=col,

        # NOTE: reduce and merge cutoff----------------------------------------
        #   Reduce will merge clusters that are similar; similarity determined
        #   by cutoff
        reduce=TRUE,

        # NOTE: color----------------------------------------------------------
        #   For more complicated designs, try the `col` argument to color the
        #   plot by another factor.
        # col="another.column",

        # NOTE: see above note
        minc=low.minc,

        plot=FALSE

        )

    n <- p$normalized
    cls <- unique(n$cluster)
    names(cls) <- cls
    cluster.counts <- sapply(cls, function (x) length(unique(n$genes[n$cluster==x])))
    clusters.keep <- names(cluster.counts)[cluster.counts > minc]
    n2 <- n[n$cluster %in% clusters.keep,]
    print(degPlotCluster(
        n2,
        time=time,
        min_genes=minc,
        col=col
        )
    )


    # In the final_clusters directory, this creates files containing lists of
    # the genes in each cluster, and adds a link to the Markdown.
    dir.create('final_clusters')
    mdcat('\n')
    for (u in unique(n2$cluster)){
        fn <- file.path('final_clusters',
                        paste0('final.', name, '.cluster.', u, '.txt'))
        write.table(unique(n2[n2$cluster==u, 'genes']), file=fn, quote=FALSE,
                    row.names=FALSE, col.names=FALSE)
        mdcat('- [', fn, '](', fn, '), cluster "', u, '" genes')
    }
    pdf.file = file.path('final_clusters', paste0(name, '.pdf'))
    dev.copy(pdf, file=pdf.file)
    dev.off()
    mdcat('- [', pdf.file, '](', pdf.file, '), PDF')

    # merge the degPattern cluster IDs with res.list
    res.list[[name]][['res']] <- add.cluster.id(clusters=n2, res=res.list[[name]][['res']], label=name)
}
```
